---
title: "00datapreparation"
author: "Noemie Coulon"
date: "27/01/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

# LOADING PACKAGES & FUNCTIONS

```{r}
if(!require(vegan)){install.packages("vegan"); library(vegan)}
if(!require(tidyverse)){install.packages("tidyverse"); library(tidyverse)}
if(!require(scales)){install.packages("scales"); library(scales)}
if(!require(raster)){install.packages("raster"); library(raster)}
if(!require(mapproj)){install.packages("mapproj"); library(mapproj)}
if(!require(sf)){install.packages("sf"); library(sf)}
if(!require(ggplot2)){install.packages("ggplot2"); library(ggplot2)}
if(!require(rnaturalearth)){install.packages("rnaturalearth"); library(rnaturalearth)}
if(!require(rnaturalearthdata)){install.packages("rnaturalearthdata"); library(rnaturalearthdata)}
if(!require(ggthemes)){install.packages("ggthemes"); library(ggthemes)}
if(!require(rgeos)){install.packages("rgeos"); library(rgeos)}

initial_wd <- "C:/Users/ncoulon/Desktop/Thèse/2021-2022/R/IV-MODEL/"
world <- ne_countries(scale = "medium", returnclass = "sf")
```

# SET DIRECTORIES
```{r}
localDir = "." 
dataDir = file.path(localDir, "data") 
ModelDir = file.path(localDir, "models") 
MixingDir = file.path(localDir, "mixing")
MFDir = file.path(localDir, "model_fit")
```

# PREPARE PA DATA
## Loading data

```{r}
ab <- read.csv("data/raw/abundance.csv", header = T)
ab_pa <- cbind(ab[, 1:12], decostand(ab[, 13:27], method = "pa"))

zero.inflation <- function(ab_pa){
  zero_inflation <- sum(apply(ab_pa[, 13:ncol(ab_pa)], 2, function(x) length(x[x == 0]))) / (nrow(ab_pa) * ncol(ab_pa[, 13:ncol(ab_pa)])) * 100
  cat(paste("---- There is", round(zero_inflation), "% of absence ----" ))
  return(zero_inflation)
}

zero_inflation <- zero.inflation(ab_pa)
```

## Zero inflation per species

```{r}
zero.inflation.sp <- function(ab_pa){
  zero_inflation_sp <- data.frame(pa=(apply(ab_pa[, 13:ncol(ab_pa)], 2, function(x) length(x[x == 1])) / nrow(ab_pa)) * 100)
  zero_inflation_sp$taxon <- rownames(zero_inflation_sp)
  for(i in 1:nrow(zero_inflation_sp)){
    cat(paste("----", zero_inflation_sp$taxon[i]), "is present in", zero_inflation_sp$pa[i], "% of the hauls ---- \n")
  }
  return(zero_inflation_sp)
}

zero_inflation_sp <- zero.inflation.sp(ab_pa)

tmp<-dplyr::filter(zero_inflation_sp, pa < 2)
cat(paste("----", nrow(tmp), "species is/are not enough sampled and removed ----"))
rm(tmp)
```

Species present in less than 2% of the total hauls are removed

```{r}
zero_inflation_sp <- dplyr::filter(zero_inflation_sp, pa >= 2)

ab_pa <- cbind(ab_pa[, 1:12],
               ab_pa[, all_of(zero_inflation_sp$taxon)])

zero_inflation <- zero.inflation(ab_pa)
```

## Gear used over time
```{r}
gears.ts <- function(ab_pa) {
  gears_ts <- data.frame(Gear = NA, count = NA, percent = NA, Year = NA)
  for (i in levels(as.factor(ab_pa$Year))) {
    year_i <- dplyr::filter(ab_pa, Year == i)
    prop.gears_i <- data.frame(fct_count(year_i$Gear))
    colnames(prop.gears_i) <- c("Gear", "count")
    prop.gears_i$percent <- (prop.gears_i$count / sum(prop.gears_i$count)) * 100
    prop.gears_i$Year <- levels(as.factor(year_i$Year))
    gears_ts <- rbind(gears_ts, prop.gears_i)
  }
  gears_ts <- gears_ts[-1, ]
  return(gears_ts)
}
gear_ts <- gears.ts(ab_pa)
```
```{r}
dev.off()
p <- ggplot(aes(x = as.factor(Year), y = percent, fill = Gear), data = gear_ts) +
  geom_col() +
  scale_fill_manual(values = c("#6accc5", "#b85e12", "#bdea5a", "#625e49", "#9e1e4a", "#5d346c", "#d8b300", "#5eb963", "#d4994a", "#5585cf","#f8766d", "#00c19f")) +
  labs(x = "Years", y = "Proportion of gears used") +
  theme_bw(50) +
  theme(panel.grid = element_blank())+
  theme(axis.text.x = element_text(angle = 90))
png(paste0("./output/context/gears_ts.png"), height = 900, width = 1500)
print(p)
dev.off()
```

Only the gears used throughout the time series are kept.

```{r}
ts <- levels(as.factor(ab_pa$Year))
gears_kept<-NA

for (i in levels(as.factor(gear_ts$Gear))){
  gear_i <- dplyr::filter(gear_ts,Gear==i)
  gear_i <- dplyr::filter(gear_i, percent != 0)
  if(nrow(gear_i) == length(ts)){
    gears_kept <- c(gears_kept, levels(as.factor(gear_i$Gear)))
  }
}
gears_kept <- gears_kept[-1]

cat(paste("----", length(levels(as.factor(ab_pa$Gear)))-length(gears_kept), "hauls has been removed ----"))

ab_pa <- dplyr::filter(ab_pa, Gear %in% gears_kept)

gear_time <- gears.ts(ab_pa)


zero_inflation <- zero.inflation(ab_pa)
zero_inflation_sp <- zero.inflation.sp(ab_pa)
```
```{r}
dev.off()
p <- ggplot(aes(x = as.factor(Year), y = percent, fill = Gear), data = gear_time) +
  geom_col() +
  scale_fill_manual(values = c("#bdea5a", "#625e49", "#d4994a", "#5585cf")) +
  labs(x = "Years", y = "Proportion of gears used") +
  theme_bw(50) +
  theme(panel.grid = element_blank())+
  theme(axis.text.x = element_text(angle = 90))

png(paste0("./output/context/gears_kept_ts.png"), height = 900, width = 1500)
print(p)
dev.off()
```

## Global gear selectivity

```{r}
tmp <- data.frame(Gear = ab_pa$Gear, HaulsID = ab_pa$HaulsID, pa = rowSums(ab_pa[, 13:ncol(ab_pa)]))
gear_selectivity <- aggregate(pa ~ Gear, data = tmp, FUN = sum)
gear_selectivity$percent <- (gear_selectivity$pa / sum(gear_selectivity$pa)) * 100
gear_selectivity

gears_removed<-NA

for (i in 1:nrow(gear_selectivity)){
  gear_i <- gear_selectivity$Gear[i]
  if(gear_selectivity$percent[i] < 1){
  ab_pa <- dplyr::filter(ab_pa, Gear != all_of(gear_i))  
  gears_removed<-c(gears_removed,gear_i)
  }
}

rm(tmp)
cat(paste("----", length(gears_removed)-1, "hauls has been removed ----"))
```

## Gear selectivity per species
```{r}
gear.selectivity.sp<-function(ab_pa){
  gear_selectivity<-data.frame(Gear=NA,pa=NA,taxon=NA)
  for (i in (13:ncol(ab_pa))){
    gear_selectivity_spi<-as.data.frame(aggregate(ab_pa[,i]~Gear,ab_pa,sum))
    colnames(gear_selectivity_spi)<-c("Gear","pa")
    gear_selectivity_spi$taxon<-colnames(ab_pa[i])
    gear_selectivity<-rbind(gear_selectivity,gear_selectivity_spi)
  }
  gear_selectivity<-gear_selectivity[-1,]
  return(gear_selectivity)
}
gear_selectivity_sp<-gear.selectivity.sp(ab_pa)
```
```{r}
dev.off()
p<-ggplot(aes(x=Gear,y=taxon,fill=log(pa)), data=gear_selectivity_sp)+
  geom_tile()+
  labs(x="Gears",y="Species") +
  theme_bw(50)+
  theme(panel.grid=element_blank()) +
  scale_fill_continuous(type = "viridis")
png(paste0("./output/context/gears_selectivity.png"), height = 900, width = 1500)
print(p)
dev.off()
```


## Continuous sampling of seasons through time
```{r}
quarter.ts <- function(ab_pa) {
  ab_pa$Quarter<-as.factor(ab_pa$Quarter)
  quarter_ts <- data.frame(Quarter = NA, count = NA, percent = NA, Year = NA)
  for (i in levels(as.factor(ab_pa$Year))) {
    year_i <- dplyr::filter(ab_pa, Year == i)
    prop.quarter_i <- data.frame(fct_count(year_i$Quarter))
    colnames(prop.quarter_i) <- c("Quarter", "count")
    prop.quarter_i$percent <- (prop.quarter_i$count / sum(prop.quarter_i$count)) * 100
    prop.quarter_i$Year <- levels(as.factor(year_i$Year))
    quarter_ts <- rbind(quarter_ts, prop.quarter_i)
  }
  quarter_ts <- quarter_ts[-1, ]
  return(quarter_ts)
}
quarter_ts <- quarter.ts(ab_pa)
```
```{r}
dev.off()
p <- ggplot(aes(x = as.factor(Year), y = percent, fill = Quarter), data = quarter_ts) +
  geom_col() +
  scale_fill_manual(values = c("#6accc5", "#b85e12", "#bdea5a", "#625e49")) +
  labs(x = "Years", y = "Proportion of quarter sampled") +
  theme_bw(50) +
  theme(panel.grid = element_blank())+
  theme(axis.text.x = element_text(angle = 90))
png(paste0("./output/context/quarter_ts.png"), height = 900, width = 1500)
print(p)
dev.off()
```

Only the quarters sampled throughout the time series are kept.

```{r}
quarter_kept<-NA
ts <- seq(1997, 2020)

for (i in levels(as.factor(quarter_ts$Quarter))){
  quarter_i <- dplyr::filter(quarter_ts, Quarter == i)
  quarter_i <- dplyr::filter(quarter_i, percent != 0)
  if(nrow(quarter_i) == length(ts)){
    quarter_kept <- c(quarter_kept, levels(as.factor(quarter_i$Quarter)))
  }
}
quarter_kept <- quarter_kept[-1]

cat(paste("----", length(levels(as.factor(ab_pa$Quarter)))-length(quarter_kept), "quarters is/are not sampled throughout the time series and removed ----"))

ab_pa <- dplyr::filter(ab_pa, Quarter %in% quarter_kept)

quarter_ts <- quarter.ts(ab_pa)

zero_inflation <- zero.inflation(ab_pa)
zero_inflation_sp <- zero.inflation.sp(ab_pa)
```
```{r}
dev.off()
p <- ggplot(aes(x = as.factor(Year), y = percent, fill = Quarter), data = quarter_ts) +
  geom_col() +
  scale_fill_manual(values = c("#6accc5", "#bdea5a", "#625e49")) +
  labs(x = "Years", y ="Proportion of quarter sampled") +
  theme_bw(50) +
  theme(panel.grid = element_blank())+
  theme(axis.text.x = element_text(angle = 90))

png(paste0("./output/context/quarters_kept_ts.png"), height = 900, width = 1500)
print(p)
dev.off()
```

## Gears used in space
```{r}
map_gears <- function(data){
  gear_colab <- NULL
  for(j in 1:nrow(data)){
      if(data$Gear[j]=="BT4A"){
        gear_colab <- c(gear_colab, "#bdea5a")
      }else if(data$Gear[j]=="BT4AI"){
        gear_colab <- c(gear_colab, "#625e49") 
      }else if(data$Gear[j]=="BT8"){
        gear_colab <- c(gear_colab, "#d4994a") 
      }else if(data$Gear[j]=="GOV"){
        gear_colab <- c(gear_colab,"#5585cf")
      }
    }
    gear_col<-c("#bdea5a", "#625e49", "#d4994a","#5585cf")
    maps::map(database = "world", ylim = c(min(data$Lat) - 7, max(data$Lat) + 2), xlim = c(min(data$Lon) - 10, max(data$Lon) + 10), fill = TRUE, col = "gray")
plot(data$Lon, data$Lat, col = gear_colab, pch = 20, cex = 2, xlab = "Longitude", ylab = "Latitude", cex.lab = 3, cex.axis = 2)
maps::map(database = "world", ylim = c(min(data$Lat) - 7, max(data$Lat) + 2), xlim = c(min(data$Lon) - 10, max(data$Lon) + 10), fill = TRUE, col = "gray", add = T)
scalebar(d = 200, xy = c(min(data$Lon), min(data$Lat)), type = "bar", below = "km", lwd = 5, divs = 2, col = "black", cex = 2, lonlat = T)
legend(max(data$Lon) - 4, max(data$Lat),
  bg = "#FFFFFF",
  legend = c("BT4A", "BT4AI", "BT8", "GOV"),
  fill = gear_col, cex = 2, title = "Gears"
)
}

dev.off()
png(paste0("./output/context/gears_map.png"), height = 900, width = 1500)
print(map_gears(ab_pa))
dev.off()
```

## Save prepared pa data
```{r}
write.csv(ab_pa,"./data/pa.csv", row.names = F)
rm(list=ls())
```

# SURVEY INVESTIGATION

## Plot surveys
```{r}
pa <- read.csv("./data/pa.csv")
levels(as.factor(pa$Survey))

Rect_ICES_27 <- read_delim("data/IFR_F_RECTANGLE_CIEM.csv", ";", escape_double = FALSE, col_types = cols(SOUTH = col_number(), 
                                                                                                   WEST = col_number(), NORTH = col_number(),
                                                                                                         EAST = col_number()), locale = locale(decimal_mark = ","),trim_ws = TRUE)
Rect_ICES_27<-dplyr::rename(Rect_ICES_27,Rect=ICESNAME)

library(RColorBrewer)
brewer.pal(9, "Set1")

dev.off()
png(paste0("./output/context/surveys_map.png"), height = 900, width = 1500)

map_surveys<-function(data){
  survey_colab <- NULL
    for(j in 1:nrow(data)){
      if(data$Survey[j]=="BTS"){
        survey_colab <- c(survey_colab,"#E41A1C")
      }else if(data$Survey[j]=="BTS-VIII"){
        survey_colab <- c(survey_colab, "#FFFF33")
      }else if(data$Survey[j]=="EVHOE"){
        survey_colab <- c(survey_colab, "#4DAF4A") 
      }else if(data$Survey[j]=="FR-CGFS"){
        survey_colab <- c(survey_colab, "#984EA3") 
      }else if(data$Survey[j]=="IE-IGFS"){
        survey_colab <- c(survey_colab, "#FF7F00")
      }else if(data$Survey[j]=="NS-IBTS"){
        survey_colab <- c(survey_colab, "#377EB8")
      }else if(data$Survey[j]=="ROCKALL"){
        survey_colab <- c(survey_colab, "#A65628")
      }else if(data$Survey[j]=="SWC-IBTS"){
        survey_colab <- c(survey_colab, "#F781BF") 
      }else if(data$Survey[j]=="SCOROC"){
        survey_colab <- c(survey_colab, "#999999") 
      }
    }
  sampling <- data.frame(Rect = levels(as.factor(data$Rect)))
  sampling <- left_join(sampling, Rect_ICES_27)
  map(database = "world", ylim = c(min(data$Lat) - 9, max(data$Lat) + 2), xlim = c(min(data$Lon) - 10, max(data$Lon) + 10), fill = TRUE, col = "gray")
  plot(data$Lon, data$Lat, col = survey_colab, pch = 20, cex = 2, xlab = "Longitude", ylab = "Latitude", cex.lab = 3, cex.axis = 3)
  survey_colab <- as.factor(survey_colab)
  survey_colab <- fct_relevel(survey_colab, c("#E41A1C", "#FFFF33", "#4DAF4A", "#984EA3", "#FF7F00", "#377EB8", "#A65628", "#F781BF", "#999999"))
  survey_colab <- as.character(levels(survey_colab))
  rect(xleft = sampling$WEST, ybottom = sampling$SOUTH, xright = sampling$EAST, ytop = sampling$NORTH)
  map(database = "world", ylim = c(min(data$Lat) - 9, max(data$Lat) + 2), xlim = c(min(data$Lon) - 10, max(data$Lon) + 10), fill = TRUE, col = "gray", add = T)
  legend(max(data$Lon) - 5, max(data$Lat) - 12,
    bg = "#FFFFFF",
    legend = c("BTS", "BTS-VIII", "EVHOE", "FR-CGFS", "IE-IGFS", "NS-IBTS", "SWC-IBTS", "ROCKALL", "SCOROC"),
    fill = survey_colab, cex = 2, title = "Surveys"
  )
  scalebar(d = 200, xy = c(min(data$Lon), min(data$Lat)), type = "bar", below = "km", lwd = 5, divs = 2, col = "black", cex = 2, lonlat = T)
}

print(map_surveys(ab_pa))
dev.off()
```

## Supplementary table

```{r}
list_surveys <- list()
ab_pa$Survey <- as.factor(ab_pa$Survey)
ab_pa$Gear <- as.factor(ab_pa$Gear)
ab_pa$Quarter <- as.factor(ab_pa$Quarter)

for (i in levels(ab_pa$Survey)){
  survey_i <- dplyr::filter(ab_pa, Survey == i)
  print(summary(survey_i))
  list_surveys <- c(list_surveys, list(survey_i))
}
```


## Survey selectivity per species
```{r}
tmp <- data.frame(Survey = ab_pa$Survey, HaulsID = ab_pa$HaulsID, pa = rowSums(ab_pa[, 13:ncol(ab_pa)]))
survey_selectivity <- aggregate(pa ~ Survey, data = tmp, FUN = sum)
survey_selectivity$percent <- (gear_selectivity$pa / sum(gear_selectivity$pa)) * 100
gear_selectivity

gears_removed<-NA

for (i in 1:nrow(gear_selectivity)){
  gear_i <- gear_selectivity$Gear[i]
  if(gear_selectivity$percent[i] < 1){
  ab_pa <- dplyr::filter(ab_pa, Gear != all_of(gear_i))  
  gears_removed<-c(gears_removed,gear_i)
  }
}
```


# REGRID PA DATA
## Loading data
```{r}
pa <- read.csv("./data/pa.csv")
baseline <- stack("./data/grid/baseline.grd")
taxon <- colnames(pa[13:ncol(pa)])
```

## Recategorize trawl gears
```{r}
pa$Gear[pa$Gear == "BT4A"] <- "BTs"
pa$Gear[pa$Gear == "BT4AI"] <- "BTs"
pa$Gear[pa$Gear == "BT8"] <- "BTs"
```


## Add gears and quarter columns
```{r}
gears <- levels(as.factor(pa$Gear))
gears

pa$GOV <- 0
pa$BTs <- 0

for (i in 1:length(gears)){
  gears_i <- gears[i]
  row_i  <- which(pa$Gear == gears_i)
  pa[row_i, gears_i] <- 1
}
```
```{r}
quarter <- levels(as.factor(pa$Quarter))
quarter

pa$Q1 <- 0
pa$Q3 <- 0
pa$Q4 <- 0

for (i in 1:length(quarter)){
  quarter_i <- paste0("Q",quarter[i])
  row_i  <- which(pa$Quarter == quarter[i])
  pa[row_i, quarter_i] <- 1
}

quarter <- c("Q1", "Q3", "Q4")
```

## Prepare the cuts
```{r}
ts <- as.numeric(levels(as.factor(pa$Year)))
pa_ts <- list()
name_pa <- NA

for (i in 1:length(ts)){
  year_i <- ts[i]
  name_pa_i <- paste0("pa_", year_i)
  pa_i <- dplyr::filter(pa, Year == year_i)
  pa_ts <- c(pa_ts, list(pa_i))
  name_pa <- c(name_pa, name_pa_i)
}

names(pa_ts) <- name_pa[-1]
```

## Rasterize pa dataset without differentiate species
```{r}
pa_ts_r <- stack()

for (i in 1:length(pa_ts)){
  pa_ts_i <- pa_ts[[i]]
  
  pa_ts_i <- data.frame(Lon = pa_ts_i$Lon, 
                    Lat = pa_ts_i$Lat, 
                    pa_ts_i[, 13:ncol(pa_ts_i)])
  pa_ts_i <- cbind(pa_ts_i[, 1:2], 
                 sum = rowSums(pa_ts_i[,3:ncol(pa_ts_i)]))
  pa_ts_i <- cbind(pa_ts_i[,1:2], 
                   Observed=decostand(pa_ts_i[,3],method="pa"))
  
  pa_ts_i_r <- rasterize(x = pa_ts_i[, c("Lon", "Lat")], 
                      y = baseline,
                      field = pa_ts_i[, "Observed"],
                      fun = function(x, ...) sum(x, ...))

  try(dev.off())  
  png(paste0("./output/context/pa_year/all_sp",names(pa_ts[i]),".png"), height = 900, width = 1200)
  print(plot(pa_ts_i_r, col = c("#FEF0D9", "#FDCC8A", "#FC8D59", "#E34A33", "#B30000")))
  dev.off()
  
  pa_ts_r <- stack(pa_ts_r, pa_ts_i_r)
}
```
```{r}
names(pa_ts_r) <- name_pa[-1]
writeRaster(pa_ts_r, "./data/grid/pa_ts_r_new", overwrite = T)
```

## Rasterize pa dataset per species/gear/quarter
```{r}
pa_sp_ts_r <- stack()

for (i in 1:length(pa_ts)){
  pa_ts_i <- pa_ts[[i]]
  for (j in 13:ncol(pa_ts_i)){
    
    sp_j <- colnames(pa_ts_i[j])
    pa_ts_sp_j <- data.frame(Lon = pa_ts_i$Lon, 
                    Lat = pa_ts_i$Lat, 
                    Observed = pa_ts_i[, j])
    pa_ts_sp_j_r <- rasterize(x = pa_ts_sp_j[, c("Lon", "Lat")], 
                      y = baseline,
                      field = pa_ts_sp_j[, "Observed"],
                      fun = function(x, ...) sum(x, ...))
    names(pa_ts_sp_j_r) <- paste0(names(pa_ts[i]), "_", sp_j)
    
    #try(dev.off()) 
    #png(paste0(initial_wd, "output/context/pa_year/",  sp_j, "_", names(pa_ts[i]), ".png"), height = 900, width = 1200)
    #print(plot(pa_ts_sp_j_r, col = c("#FEF0D9", "#FDCC8A", "#FC8D59", "#E34A33", "#B30000")))
    #dev.off()
  
  pa_sp_ts_r <- stack(pa_sp_ts_r, pa_ts_sp_j_r)
  }
}

writeRaster(pa_sp_ts_r, "./data/grid/pa_year_sp_gears_quarter_r_new", overwrite = T)
```

# REMAKE A DATAFRAME
## One dataframe per year
```{r}
pa_sp_ts_r <- stack("./data/grid/pa_year_sp_gears_quarter_r_new")

n_col <- as.numeric(length(pa[,13:ncol(pa)]))
start_year <- seq(1, nlayers(pa_sp_ts_r), n_col)
df <- dplyr::select(pa, Lon, Lat, Year, Quarter, all_of(taxon))
df[,] <- NA
df <- na.omit(df)

for (i in 1:length(start_year)){
  ts_i <- ts[i]
  start <- start_year[i]
  end <- start_year[i] + n_col
  seq_i <- seq(start, end)
  seq_i <- seq_i[-length(seq_i)]
  df_i <- as.data.frame(pa_sp_ts_r[[seq_i]], na.rm = T, xy = T)
  df_i <- cbind(df_i[,1:2], decostand(df_i[,3:ncol(df_i)], method = "pa"))
  df_i$Year <- ts_i
  df_i <- relocate(df_i, Year, .after = y)
  colnames(df_i) <- c("Lon", "Lat", "Year", taxon, gears, quarter)
  tmp <- dplyr::select(df_i, all_of(c(gears,quarter)))
  tmp[tmp == 0] <- NA
  tmp["BTs"][tmp["BTs"] == 1]  <- "BTs"
  tmp["GOV"][tmp["GOV"] == 1] <- "GOV"
  tmp["Q1"][tmp["Q1"] == 1] <- "Q1"
  tmp["Q3"][tmp["Q3"] == 1] <- "Q3"
  tmp["Q4"][tmp["Q4"] == 1] <- "Q4"
  df_i <- cbind(df_i[1:(3+length(taxon))], tmp)
  df_i <- unite(df_i, col = Gear, all_of(gears), sep = "_", remove = T, na.rm = T)
  df_i <- relocate(df_i, Gear, .after = Year)
  df_i <- unite(df_i, col = Quarter, all_of(quarter), sep = "_", remove = T, na.rm = T)
  df_i <- relocate(df_i, Quarter, .after = Gear)
  df <- rbind(df, df_i)
}

write.csv(df, "./data/pa_grid_new.csv", row.names = F)
```

## Global zero inflation
```{r}
df <- read.csv("./data/pa_grid_new.csv")

zero.inflation <- function(df){
  zero_inflation <- sum(apply(df[, 6:ncol(df)], 2, function(x) length(x[x == 0]))) / (nrow(df) * ncol(df[, 6:ncol(df)])) * 100
  cat(paste("---- There is", round(zero_inflation), "% of absence ----" ))
  return(zero_inflation)
}

zero_inflation <- zero.inflation(df)
```

## Zero inflation per species
```{r}
zero.inflation.sp <- function(df){
  zero_inflation_sp <- data.frame(pa=(apply(df[, 6:ncol(df)], 2, function(x) length(x[x == 1])) / nrow(df)) * 100)
  zero_inflation_sp$taxon <- rownames(zero_inflation_sp)
  for(i in 1:nrow(zero_inflation_sp)){
    cat(paste("----", zero_inflation_sp$taxon[i]), "is present in", zero_inflation_sp$pa[i], "% of the hauls ---- \n")
  }
  return(zero_inflation_sp)
}

zero_inflation_sp <- zero.inflation.sp(df)
```

## Area covered by species
```{r}
df <- unite(df, col = id, Lon, Lat, remove = F)
tmp <- aggregate(df[7:ncol(df)], by = list(df$id), FUN = sum)

for (i in 2:ncol(tmp)){
  tmp_i<- tmp[i]
  colnames(tmp_i) <- c("obs")
  tmp_i <- dplyr::filter(tmp_i, obs !=0)
  zero_inflation_sp$obs_surface[i-1] <- nrow(tmp_i)/nrow(tmp)*100
}

zero_inflation_sp
```

## Dispersion per species
```{r}
tmp <- separate(tmp, Group.1, c("Lon", "Lat"), remove = F, sep = "_")
tmp$Lon <- as.numeric(tmp$Lon)
tmp$Lat <- as.numeric(tmp$Lat)

for (i in 4:ncol(tmp)){
  tmp_i<- cbind(tmp[2:3], tmp[i])
  colnames(tmp_i) <- c("Lon", "Lat", "obs")
  tmp_i <- dplyr::filter(tmp_i, obs !=0)
  zero_inflation_sp$disp_lat[i-3] <- sd(tmp_i$Lat)
  zero_inflation_sp$disp_lon[i-3] <- sd(tmp_i$Lon)
}
zero_inflation_sp
```

## Map of total species presence
```{r}
data <- read.csv("./data/pa_grid_new.csv", header = T)
data$Lon <- round(data$Lon, digits = 5)
data$Lat <- round(data$Lat, digits = 5)
data <- unite(data, col = id, Lon, Lat, remove = F)

grid <- read.csv(paste0(dataDir, "/grid_projection.csv"))
grid <- grid[, -1]
grid <- dplyr::filter(grid, Year == 1997)
```
```{r}
for (i in 7:ncol(data)){
  data_i <- cbind(data[, 1:5], sum = data[, i])
  
  sum_i <- aggregate(sum ~ id, data = data_i, FUN = sum)
  sum_i <- separate(sum_i, id, c("Lon", "Lat"), remove = F, sep = "_")
  sum_i$Lon <- as.numeric(sum_i$Lon)
  sum_i$Lat <- as.numeric(sum_i$Lat)
  
  map <- ggplot(world) +
  geom_tile(data = sum_i, 
                 aes(x = Lon, y = Lat, 
                 fill = sum)) +
      geom_sf(data = world, 
             fill = alpha("lightgrey", 1)) +
      coord_sf(xlim = c(min(grid$Lon), max(grid$Lon)), 
           ylim = c(min(grid$Lat), max(grid$Lat))) +
      theme_bw(150) +
      theme(panel.grid = element_blank()) +
    theme(axis.title = element_blank()) +
    theme(legend.position = "bottom", legend.title = element_blank()) +
    labs(fill = colnames(data[i])) +
    scale_fill_gradient2(low = "#343d46", high = "#ff084a", midpoint = 1) +
    guides(fill = guide_colourbar(barwidth = 50,
                                barheight = 2))
  
  try(dev.off())  
  png(paste0("./output/context/presence/", colnames(data[i]), "_presence.png"), width = 2100, height = 3000)
    print(map)
    dev.off()
}
```