---
title: "05_predictions_baseline"
author: "Noemie Coulon"
date: "28/02/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# LOADING PACKAGES & FUNCTIONS
```{r}
if(!require(tidyverse)){install.packages("tidyverse"); library(tidyverse)}
if(!require(knitr)){install.packages("knitr"); library(knitr)}
if(!require(Hmsc)){install.packages("Hmsc"); library(Hmsc)}
if(!require(abind)){install.packages("abind"); library(abind)}
if(!require(ggplot2)){install.packages("ggplot2"); library(ggplot2)}
if(!require(sf)){install.packages("sf"); library(sf)}
sf::sf_use_s2(FALSE) 
if(!require(rnaturalearth)){install.packages("rnaturalearth"); library(rnaturalearth)}
if(!require(ggthemes)){install.packages("ggthemes"); library(ggthemes)}
if(!require(rgeos)){install.packages("rgeos"); library(rgeos)}
if(!require(factoextra)){install.packages("factoextra"); library(factoextra)}
if(!require(FactoMineR)){install.packages("FactoMineR"); library(FactoMineR)}
if(!require(sp)){install.packages("sp"); library(sp)}
if(!require(prevR)){install.packages("prevR"); library(prevR)}
if(!require(raster)){install.packages("raster"); library(raster)}
if(!require(trend)){install.packages("trend"); library(trend)}
if(!require(viridis)){install.packages("viridis"); library(viridis)}
if(!require(GDAtools)){install.packages("GDAtools"); library(GDAtools)}

world <- ne_countries(scale = "medium", returnclass = "sf")
```

# SET DIRECTORIES
```{r}
localDir = "." 
dataDir = file.path(localDir, "data") 
ModelDir = file.path(localDir, "models") 
MixingDir = file.path(localDir, "mixing")
MFDir = file.path(localDir, "model_fit")
```

# LOAD MODEL, MODEL HINDCAST and BASELINE
```{r}
load(paste0(ModelDir, "/pa_traits_phylo_probit_thin_3.RData"))

pa_hindcast <- read.csv(paste0(dataDir, "/hindcast/pa_hindcast_traits.csv"))
envt_hindcast <- read.csv(paste0(dataDir, "/hindcast/envt_hindcast_traits.csv"))
envt_hindcast <- dplyr::select(envt_hindcast, -substrate)

grid <- read.csv(paste0(dataDir, "/envt_var_selected_df.csv"))
grid$depth <- abs(grid$depth)
grid <- dplyr::select(grid, -substrate)
```

# SPATIAL VARIATION 

## Do not extrapolate

PCA is performed to resume environmental conditions

```{r}
grid_pca <- unite(grid, id, c("id", "Year"), sep= "_", remove = F)
grid_pca <- dplyr::filter(grid_pca, ! id %in% envt_hindcast$id)
grid_pca$proj <- "forecast"

envt_hindcast_pca <- envt_hindcast
envt_hindcast_pca$proj <- "hindcast"
envt_hindcast_pca <- relocate(envt_hindcast_pca, id, .before = Lon)

grid_pca <- rbind(grid_pca, envt_hindcast_pca)
drivers <- grid_pca[, 5:10]
```
```{r}
res.pca <- PCA(grid_pca[, 5:10], axes = c(1, 2))
coord_pca <- cbind(grid_pca, res.pca$ind$coord)
```
```{r}
fviz_pca_ind (res.pca, col.ind = grid_pca$proj, geom = "point", axes = c(1, 2))
fviz_eig(res.pca, addlabels = T)
```
```{r}
fviz_contrib(res.pca, choice = "var", axes = 1:2, top = 10)
```

### Convex hull

The convex hull of hindcast data is calculated

```{r}
conv.hull <- coord_pca %>%
  group_by(proj) %>%
  mutate(hull = 1:n(), hull = factor(hull, chull(Dim.1, Dim.2))) %>%
  arrange(hull)
```
```{r}
ggplot(conv.hull, aes(Dim.1, Dim.2, color = proj, fill = proj)) +
  geom_point() +
  geom_polygon(data = filter(conv.hull, !is.na(hull)), alpha = 0.4) +
  guides(fill = FALSE, colour = FALSE) +
  theme_bw(60) +
  theme(panel.grid = element_blank())
```

We would like to know which environmental condition is within/without the hindcast convex hull

```{r}
poly_hull <- sp::Polygon(cbind(conv.hull[, 12], conv.hull[, 13]))
poly_hull <- sp::Polygons(list(poly_hull), ID = "hindcast_hull")
poly_hull <- sp::SpatialPolygons(list(poly_hull))

coord_pca$place <- point.in.SpatialPolygons(coord_pca$Dim.1, coord_pca$Dim.2, poly_hull)
coord_pca$place[coord_pca$place == "TRUE"] <- "in"
coord_pca$place[coord_pca$place == "FALSE"] <- "out"
```
```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")

ggplot(world) +
  geom_tile(data = coord_pca, 
                 aes(x = Lon, y = Lat, 
                 fill = place), 
                 size = .85) +
      geom_sf(data = world, 
              fill = alpha("lightgrey", 1)) +
      coord_sf(xlim = c(min(coord_pca$Lon), max(coord_pca$Lon)), 
           ylim = c(min(coord_pca$Lat), max(coord_pca$Lat))) +
      theme_bw(15) +
      theme(panel.grid = element_blank()) +
      labs( fill = "Hindcast\nconvex hull")+ 
      xlab("Longitude") + 
      ylab("Latitude")
```
```{r}
coord_pca <- dplyr::filter(coord_pca, place == "in")
coord_pca<- unite(coord_pca, id, c("Lon", "Lat"), sep = "_", remove = T)

write.csv(coord_pca, paste0(dataDir, "/coord_pca.csv"))
```

### Filter
```{r}
min.depth <- max(envt_hindcast$depth)
max.depth <- min(envt_hindcast$depth)

grid <- dplyr::filter(grid, depth < min.depth)
grid <- dplyr::filter(grid, depth > max.depth)

grid <- dplyr::filter(grid, Lon > -12)

supp <- rbind(dplyr::filter(grid, Lon < -3 & Lat > 60.5),
              dplyr::filter(grid, Lon > 4 & Lat > 58))
grid <- dplyr::filter(grid, !id  %in% supp$id)

write.csv(grid, paste0(dataDir, "/grid_projection.csv"))
```
```{r}
ggplot(world) +
  geom_tile(data = grid, 
                 aes(x = Lon, y = Lat, 
                 fill = depth), 
                 size = .85) +
      geom_sf(data = world, 
              fill = alpha("lightgrey", 1)) +
      coord_sf(xlim = c(min(grid$Lon), max(grid$Lon)), 
           ylim = c(min(grid$Lat), max(grid$Lat))) +
      theme_bw(15) +
      theme(panel.grid = element_blank()) +
      labs( fill = "Depth")+ 
      xlab("Longitude") + 
      ylab("Latitude")
```

## Predictions

### Context
```{r}
grid <- read.csv(paste0(dataDir, "/grid_projection.csv"))
grid <- grid[, -1]
ts <- levels(as.factor(envt_hindcast$Year))
```
```{r}
xy.grid <- as.matrix(cbind(grid$Lon, grid$Lat))
XData.grid <- dplyr::select(grid, -Lon, -Lat)

EpredY_list <- list()

try(dir.create(paste0(MFDir,"/mean_predictions"), recursive = T))

for (i in 1:length(ts)){
  year_i <- as.numeric(ts[i])
  
  cat(paste("----", Sys.time(), "----", year_i, "projections started ----\n", sep = " "))
  
  grid_i <- dplyr::filter(grid, Year == all_of(year_i))
  xy.grid <- as.matrix(cbind(grid_i$Lon, grid_i$Lat))
  XData.grid <- dplyr::select(grid_i, -Lon, -Lat)
  
  Gradient <- prepareGradient(m, XDataNew = XData.grid, sDataNew = list(Quarter = xy.grid, Gear = xy.grid))
  predY <- predict(m, Gradient = Gradient, predictEtaMean = T, type = "response")
  
  save(predY, file = paste0(MFDir,"/all_predictions/predY_", year_i, "_list_poly_traits_3.Rdata"))
  EpredY <- Reduce("+", predY) / length(predY)
  
  EpredY_list <- c(EpredY_list, list(EpredY))
}

save(EpredY_list, file = paste0(MFDir,"/mean_predictions/EpredY_list_traits_phylo_3.Rdata"))
```

## Maps
```{r}
load(paste0(MFDir,"/mean_predictions/EpredY_list_traits_phylo_3.Rdata"))
world <- ne_countries(scale = "medium", returnclass = "sf")

for (i in 1:length(ts)){
  year_i <- as.numeric(ts[i])
  
  cat(paste("----", Sys.time(), "----", year_i, "projections started ----\n", sep = " "))
  
  EpredY_i <- EpredY_list[[i]]
  pa_hindcast_i <- dplyr::filter(pa_hindcast, Year == all_of(year_i))
  
  sp_tot <- ncol(EpredY_i)
  sp_proj <- 0
  
  grid_i <- dplyr::filter(grid, Year == all_of(year_i))
  xy.grid <- as.matrix(cbind(grid_i$Lon, grid_i$Lat))
  XData.grid <- dplyr::select(grid_i, -Lon, -Lat)
  
  for (j in 7:ncol(pa_hindcast_i)){
    
    taxon_j <- colnames(pa_hindcast_i[j])
    
    pa_hindcast_j <- dplyr:: select(pa_hindcast_i, Lon, Lat, all_of(taxon_j))
    indexNames <- which(pa_hindcast_j[3] == 0)
    pa_hindcast_j <- pa_hindcast_j[-indexNames,]
    
    pred_sp <- EpredY_i[, taxon_j]
    mapData <- data.frame(xy.grid, pred_sp)
    colnames(mapData) <- c("Lon", "Lat", "pred_sp")
    sp_j <- sub("_", " ", taxon_j)
    
    pred_export <- mapData
    pred_export$taxon <- taxon_j
    pred_export$Year <- year_i
    
    try(dir.create(paste0(dataDir, "/grid/predictions")))
    write.csv(pred_export, paste0(dataDir, "/grid/predictions/df_suitability_", taxon_j, "_", year_i, ".csv"), row.names = F)
    
    
    mapData_raster <- rasterFromXYZ(mapData, digits = 3)
    names(mapData_raster) <- paste0("suitability_poly_traits_3_", taxon_j, "_", year_i)
    crs(mapData_raster) <- "+proj=longlat +datum=WGS84"
    try(dir.create(paste0(dataDir, "/grid/predictions/raster")))
    writeRaster(mapData_raster, paste0(dataDir, "/grid/predictions/raster/suitability_poly_traits_3bis_", taxon_j, "_", year_i),
                overwrite = TRUE)
    
    sp <- ggplot(world)  +
      geom_tile(data = mapData, 
                  aes(x = Lon, y = Lat, fill  = pred_sp)) +
      geom_point(data = pa_hindcast_j, 
                 aes(x = Lon, y = Lat), 
                 col = "black", 
                 size = 2) +
      geom_sf(data = world, 
              fill = alpha("lightgrey", 1)) +
      coord_sf(xlim = c(min(mapData$Lon), max(mapData$Lon)), 
           ylim = c(min(mapData$Lat), max(mapData$Lat))) +
      theme_bw(100) +
      theme(panel.grid = element_blank(),
            axis.title = element_blank(),
            axis.ticks = element_blank(),
            axis.text = element_blank(),
            legend.position = "bottom") +
      labs(title = paste(sp_j, "-", year_i, sep = " "), 
           fill = "Habitat\nsuitability   ") +
      scale_fill_gradientn(colors = c("#440154FF", "#39568CFF", "#1F968BFF", "#73D055FF", "#FDE725FF"), 
                       limits = c(0, 1)) +
      guides(fill = guide_colorbar(barwidth = 65,
                                   barheight = 3))
    
    try(dir.create(paste0(MFDir,"/", taxon_j), recursive = T))
    
    try(dev.off())  
    png(paste0(MFDir,"/", taxon_j, "/", taxon_j, "_", year_i, "_poly_traits_3bis.png"), height = 3000, width = 2100)
    print(sp)
    dev.off()
    
    try(dev.off())  
    png(paste0(MFDir,"/GIFs/", taxon_j, "_", year_i, "_poly_traits_3bis.png"), height = 3000, width = 2100)
    print(sp)
    dev.off()
    
    sp_proj <- sp_proj + 1
    
    cat(paste( "----", sp_proj, "/", sp_tot, "species projected ----\n", sep = " "))
  }
}
```

# SPECIES RICHNESS, RCP, CWM

## Data
```{r}
load(paste0(MFDir,"/mean_predictions/EpredY_list_traits_phylo_3.Rdata"))
world <- ne_countries(scale = "medium", returnclass = "sf")
grid <- read.csv(paste0(dataDir, "/grid_projection.csv"))
grid <- grid[, -1]
ts <- levels(as.factor(grid$Year))
```

## Maps
```{r}
for (i in 1:length(ts)){
  year_i <- as.numeric(ts[i])
  grid_i <- dplyr::filter(grid, Year == year_i)
  
  cat(paste("----", Sys.time(), "----", year_i, "species richness projections started ----\n", sep = " "))
  
  
  EpredY_i <- EpredY_list[[i]]
  S_i <- rowSums(EpredY_i)
  predT_i <- (EpredY_i%*%m$Tr) / matrix(rep(S_i, m$nt), ncol = m$nt) 
  
  RCP_i <- kmeans(EpredY_i, 4)
  MCS <- matrix(NA, m$ns, 4)
  for (i in 1:4){
    MCS[, i] <- names(rev(sort(RCP_i$centers[i, ])))
    }
  colnames(MCS) = c("Region 1","Region 2","Region 3", "Region 4")
  MCS[1:3, ]
  
  mapData_i <- data.frame(grid_i[, 2:5], S_i, predT_i, RCP_i$cluster)
  
  try(dir.create(paste0(dataDir, "/grid/predictions")))
  write.csv(mapData_i, paste0(dataDir, "/grid/predictions/community/df_S_RCP_CWM_", year_i, ".csv"), row.names = F)
    
    S <- ggplot(world)  +
      geom_tile(data = mapData_i, 
                  aes(x = Lon, y = Lat, fill  = S_i)) +
      geom_sf(data = world, 
              fill = alpha("lightgrey", 1)) +
      coord_sf(xlim = c(min(mapData_i$Lon), max(mapData_i$Lon)), 
           ylim = c(min(mapData_i$Lat), max(mapData_i$Lat))) +
      theme_bw(30) +
      theme(panel.grid = element_blank()) +
      labs(title = paste("Predicted species richness -", year_i, sep = " "), 
           fill = "Species richness    ")+ 
      xlab("Longitude") + 
      ylab("Latitude") +
      scale_fill_gradientn(colors = c("#440154FF", "#39568CFF", "#1F968BFF", "#73D055FF", "#FDE725FF"), 
                       limits = c(0, 6)) +
      theme(legend.position = "bottom") +
      guides(fill = guide_colorbar(barwidth = 30))
    
    try(dir.create(paste0(MFDir,"/species_richness"), recursive = T))
    
    try(dev.off())  
    png(paste0(MFDir,"/species_richness/S_", year_i, "_poly_traits_3.png"), height = 1200, width = 900)
    print(S)
    dev.off()
    
    RCP <- ggplot(world)  +
      geom_tile(data = mapData_i, 
                  aes(x = Lon, y = Lat, fill  = as.factor(RCP_i.cluster))) +
      geom_sf(data = world, 
              fill = alpha("lightgrey", 1)) +
      coord_sf(xlim = c(min(mapData_i$Lon), max(mapData_i$Lon)), 
           ylim = c(min(mapData_i$Lat), max(mapData_i$Lat))) +
      theme_bw(30) +
      theme(panel.grid = element_blank()) +
      labs(title = paste("Region of common profil -", year_i, sep = " "), 
           fill = "Region of common profil    ")+ 
      xlab("Longitude") + 
      ylab("Latitude") +
      scale_fill_brewer(palette = 1) +
      theme(legend.position = "bottom") 
    
    try(dir.create(paste0(MFDir,"/region_of_common_profil"), recursive = T))
    
    try(dev.off())  
    png(paste0(MFDir,"/region_of_common_profil/RCP_", year_i, "_poly_traits_3.png"), height = 1200, width = 900)
    print(RCP)
    dev.off()
    
     tl <- ggplot(world)  +
      geom_tile(data = mapData_i, 
                  aes(x = Lon, y = Lat, fill  = tl)) +
      geom_sf(data = world, 
              fill = alpha("lightgrey", 1)) +
      coord_sf(xlim = c(min(mapData_i$Lon), max(mapData_i$Lon)), 
           ylim = c(min(mapData_i$Lat), max(mapData_i$Lat))) +
      theme_bw(30) +
      theme(panel.grid = element_blank()) +
      labs(title = paste("Predicted CWM for trophic level -", year_i, sep = " "), 
           fill = "CWM - trophic level    ")+ 
      xlab("Longitude") + 
      ylab("Latitude") +
      scale_fill_gradientn(colors = c("#440154FF", "#39568CFF", "#1F968BFF", "#73D055FF", "#FDE725FF"), 
                       limits = c(3.75, 4.01)) +
      theme(legend.position = "bottom") +
      guides(fill = guide_colorbar(barwidth = 30))
    
    try(dir.create(paste0(MFDir,"/CWM"), recursive = T))
    
    try(dev.off())  
    png(paste0(MFDir,"/CWM/CWM_trophic_level_", year_i, "_poly_traits_3.png"), height = 1200, width = 900)
    print(tl)
    
    fecundity <- ggplot(world)  +
      geom_tile(data = mapData_i, 
                  aes(x = Lon, y = Lat, fill  = fecundity)) +
      geom_sf(data = world, 
              fill = alpha("lightgrey", 1)) +
      coord_sf(xlim = c(min(mapData_i$Lon), max(mapData_i$Lon)), 
           ylim = c(min(mapData_i$Lat), max(mapData_i$Lat))) +
      theme_bw(30) +
      theme(panel.grid = element_blank()) +
      labs(title = paste("Predicted CWM for fecundity -", year_i, sep = " "), 
           fill = "CWM - fecundity    ")+ 
      xlab("Longitude") + 
      ylab("Latitude") +
      scale_fill_gradientn(colors = c("#440154FF", "#39568CFF", "#1F968BFF", "#73D055FF", "#FDE725FF"), 
                       limits = c(51, 88)) +
      theme(legend.position = "bottom") +
      guides(fill = guide_colorbar(barwidth = 30))
    
    try(dev.off())  
    png(paste0(MFDir,"/CWM/CWM_fecundity_", year_i, "_poly_traits_3.png"), height = 1200, width = 900)
    print(fecundity)
    
    Lm <- ggplot(world)  +
      geom_tile(data = mapData_i, 
                  aes(x = Lon, y = Lat, fill  = length.maturity)) +
      geom_sf(data = world, 
              fill = alpha("lightgrey", 1)) +
      coord_sf(xlim = c(min(mapData_i$Lon), max(mapData_i$Lon)), 
           ylim = c(min(mapData_i$Lat), max(mapData_i$Lat))) +
      theme_bw(30) +
      theme(panel.grid = element_blank()) +
      labs(title = paste("Predicted CWM for length at maturity -", year_i, sep = " "), 
           fill = "CWM - length at maturity    ")+ 
      xlab("Longitude") + 
      ylab("Latitude") +
      scale_fill_gradientn(colors = c("#440154FF", "#39568CFF", "#1F968BFF", "#73D055FF", "#FDE725FF"), 
                       limits = c(57, 71)) +
      theme(legend.position = "bottom") +
      guides(fill = guide_colorbar(barwidth = 30))
    
    try(dev.off())  
    png(paste0(MFDir,"/CWM/CWM_length_at_maturity_", year_i, "_poly_traits_3.png"), height = 1200, width = 900)
    print(Lm)
    
    Lmax <- ggplot(world)  +
      geom_tile(data = mapData_i, 
                  aes(x = Lon, y = Lat, fill  = length.max)) +
      geom_sf(data = world, 
              fill = alpha("lightgrey", 1)) +
      coord_sf(xlim = c(min(mapData_i$Lon), max(mapData_i$Lon)), 
           ylim = c(min(mapData_i$Lat), max(mapData_i$Lat))) +
      theme_bw(30) +
      theme(panel.grid = element_blank()) +
      labs(title = paste("Predicted CWM for maximal length -", year_i, sep = " "), 
           fill = "CWM - maximal length    ")+ 
      xlab("Longitude") + 
      ylab("Latitude") +
      scale_fill_gradientn(colors = c("#440154FF", "#39568CFF", "#1F968BFF", "#73D055FF", "#FDE725FF"), 
                       limits = c(76, 111)) +
      theme(legend.position = "bottom") +
      guides(fill = guide_colorbar(barwidth = 30))
    
    try(dev.off())  
    png(paste0(MFDir,"/CWM/CWM_maximal_length_", year_i, "_poly_traits_3.png"), height = 1200, width = 900)
    print(Lmax)
    
    try(dev.off)
}
```