---
title: "06_change_in_distribution_area"
output: html_document
date: '2022-03-07'
editor_options: 
  chunk_output_type: console
---

# LOADING PACKAGE
```{r}
if(!require(tidyverse)){install.packages("tidyverse"); library(tidyverse)}
if(!require(raster)){install.packages("raster"); library(raster)}
if(!require(ncdf4)){install.packages("ncdf4"); library(ncdf4)}
if(!require(Hmsc)){install.packages("Hmsc"); library(Hmsc)}
if(!require(sf)){install.packages("sf"); library(sf)}
if(!require(ggplot2)){install.packages("ggplot2"); library(ggplot2)}
if(!require(ggalluvial)){install.packages("ggalluvial"); library(ggalluvial)}
if(!require(rnaturalearth)){install.packages("rnaturalearth"); library(rnaturalearth)}
if(!require(rnaturalearthdata)){install.packages("rnaturalearthdata"); library(rnaturalearthdata)}
if(!require(ggthemes)){install.packages("ggthemes"); library(ggthemes)}
if(!require(rgeos)){install.packages("rgeos"); library(rgeos)}
if(!require(Kendall)){install.packages("Kendall"); library(Kendall)}
if(!require(viridis)){install.packages("viridis"); library(viridis)}
if(!require(plotly)){install.packages("plotly"); library(plotly)}
if(!require(Hmisc)){install.packages("Hmisc"); library(Hmisc)}
if(!require(factoextra)){install.packages("factoextra"); library(factoextra)}
if(!require(FactoMineR)){install.packages("FactoMineR"); library(FactoMineR)}
if(!require(trend)){install.packages("trend"); library(trend)}

world <- ne_countries(scale = "medium", returnclass = "sf")
```

# SET DIRECTORIES
```{r}
localDir = "." 
dataDir = file.path(localDir, "data") 
ModelDir = file.path(localDir, "models") 
MixingDir = file.path(localDir, "mixing")
MFDir = file.path(localDir, "model_fit")
```

# LOADING DATA
## Grid
```{r}
grid <- read.csv(paste0(dataDir, "/grid_projection.csv"))
grid <- grid[, -1]

grid_raster <- raster(paste0(dataDir, "/grid/baseline.grd"))
area_raster <- area(grid_raster)
area_raster_df <- as.data.frame(area_raster, xy = T)
colnames(area_raster_df) <- c("Lon", "Lat", "area")

area_raster_df$Lon <- round(area_raster_df$Lon, digits = 5)
area_raster_df$Lat <- round(area_raster_df$Lat, digits = 5)
grid$Lon <- round(grid$Lon, digits = 5)
grid$Lat <- round(grid$Lat, digits = 5)

grid <- merge(grid, area_raster_df)
grid$depth <- abs(grid$depth)
```
```{r}
ggplot(world) +
  geom_tile(data = grid, 
                 aes(x = Lon, y = Lat, 
                 fill = depth), 
                 size = .85) +
      geom_sf(data = world, 
              fill = alpha("lightgrey", 1)) +
      coord_sf(xlim = c(min(grid$Lon), max(grid$Lon)), 
           ylim = c(min(grid$Lat), max(grid$Lat))) +
      theme_bw(15) +
      theme(panel.grid = element_blank()) +
      labs( fill = "Depth")+ 
      xlab("Longitude") + 
      ylab("Latitude")
```

## Predictions
```{r}
load(paste0(ModelDir, "/pa_traits_phylo_probit_thin_3.RData"))
load(paste0(MFDir, "/mean_predictions/EpredY_list_traits_phylo_3.Rdata"))
```

# HABITAT SUITABILITY SIZE
```{r}
weighted_area <- data.frame(taxon = NA, Year = NA, value = NA)
start <- 1997

for (i in 1:length(EpredY_list)){
  
  EpredY_i <- EpredY_list[[i]]
  grid_i <- dplyr::filter(grid, Year == start)
  EpredY_i <- data.frame(cbind(grid_i[, 1:2], area = grid_i[, 11]), EpredY_i)
  
  weighted_area_i <- (EpredY_i$area * EpredY_i[, 4:ncol(EpredY_i)]) 
  
  weighted_area <- rbind(weighted_area, 
                         data.frame(taxon = colnames(weighted_area_i), Year = start, value = colSums(weighted_area_i)))
  
  start <- start + 1 
}

weighted_area <- weighted_area[-1, ]

write.csv(weighted_area, file = paste0(MFDir, "/weighted_area_time_series_traits_3.csv"), row.names = F)

base::colSums(grid_i[, 5:11]) # total surface area
```

## Stat
```{r}
weighted_area <- read.csv(paste0(MFDir, "/weighted_area_time_series_traits_3.csv"))

excel <- weighted_area
excel$value <- round(excel$value)

excel <- pivot_wider(excel, values_from = value, names_from = taxon)
excel
write.csv(excel, paste0(MFDir, "./paper_weighted_area_time_series_traits_3.csv"), row.names = F)
```

### Sen

Estimates of the regression coefficient based on Kendall's tau

```{r}
trend_stat <- data.frame(Sen = NA, pvalue = NA, taxon = NA, change = NA)

for (i in levels(as.factor(weighted_area$taxon))){
  factor_i <- filter(weighted_area, taxon == i)
  
  try(test_i <- sens.slope(factor_i$value, conf.level = 0.95))
  trend_i <- data.frame(Sen = test_i$estimates, pvalue = test_i$p.value, taxon = i)
  
  if(trend_i$pvalue < 0.05 & trend_i$Sen > 0){
    trend_i$change <- "Increasing"
  }else if(trend_i$pvalue < 0.05 & trend_i$Sen < 0){
    trend_i$change <- "Decreasing"
  }else{
    trend_i$change <- "Unchanged"
  }
  
  trend_stat <- rbind(trend_stat, trend_i)
}

trend_stat <- trend_stat[-1, ]
trend_stat

write.csv(trend_stat, paste0(MFDir, "/weighted_area_trend_stat_traits_3.csv"), row.names = F)
```

## Plots
```{r}
weighted_area <- read.csv(paste0(MFDir, "/weighted_area_time_series_traits_3.csv"))
weighted_area_stat <- read.csv(paste0(MFDir, "/weighted_area_trend_stat_traits_3.csv"))
```
```{r}
ggplot(aes(x = Year, y = value), data = weighted_area) +
  geom_point(aes(color = taxon)) +
  scale_color_manual(values = c("black", "black", "red", "red", "red", "black", "black", "red", "black")) +
  scale_fill_manual(values = c("black", "black", "red", "red", "red", "black", "black", "red", "black")) +
  theme_bw(15)+
  labs(col = NULL, y = "Weighted area (km2)", x = NULL) +
  theme(panel.grid = element_blank()) +
  theme(legend.position = "bottom") +
  geom_smooth(aes(color = taxon, fill = taxon), method = "lm") 
```
```{r}
weighted_area_stat$time_series <- (weighted_area_stat$Sen * 24) / 1051523.165
```
```{r}
area <- ggplot() +
  geom_col(aes (y = taxon, x = Sen, fill = taxon), data = weighted_area_stat) +
  theme_bw(150) +
  theme(panel.grid = element_blank()) +
  scale_y_discrete(limits = c("Squalus_acanthias",
                              "Scyliorhinus_stellaris",
                              "Scyliorhinus_canicula",
                              "Raja_montagui",
                              "Raja_clavata",
                              "Raja_brachyura",
                              "Mustelus_asterias",
                              "Leucoraja_naevus",
                              "Amblyraja_radiata")) +
  scale_fill_manual(values = c("#0d2fb1ff","#999999ff", "#e69f00ff", "#56b4e9ff","#009e73ff", "#fff400ff","#0085f2ff","#e98377ff","#ff71fcff")) +
  guides(fill = "none")
  

try(dev.off())  
  png(paste0("./output/traits_phylo_3/area_change.png"), height = 1500, width = 3000)
    print(area)
    dev.off() 
```



# PRESENCE PROBABILITY IN SPACE

## Dataframe remaking
```{r}
list_csv <- list.files(paste0(dataDir, "/grid/predictions"), pattern = ".csv", full.names = TRUE)
all_csv <- lapply(list_csv, readr::read_csv)
df_csv <- dplyr::bind_rows(all_csv)
write.csv(df_csv, file = paste0(MFDir, "/df_preds_traits_3.csv"), row.names = F)
```

## Trends
```{r}
df_predY <- read.csv(paste0(MFDir, "/df_preds_traits_3.csv"))
df_predY <- unite(df_predY, id, c("Lon", "Lat"), remove = F)

trend_cell <- data.frame(id = NA, Sen = NA, pvalue = NA, taxon = NA)

for (i in levels(as.factor(df_predY$taxon))){
  
   taxon_i <- i
   cat(paste("----", Sys.time(), "----", taxon_i, "trends research began ----\n", sep = " "))
   predY_i <- dplyr::filter(df_predY, taxon == i)
   
   for (j in levels(as.factor(predY_i$id))){
     
     cell_j <- dplyr::filter(predY_i, id == j)
     
     try(test_i <- sens.slope(cell_j$pred_sp, conf.level = 0.95))
     
     trend_i <- data.frame(Sen = test_i$estimates, pvalue = test_i$p.value, taxon = taxon_i)
     
     if(trend_i$pvalue == "NaN"){
       trend_cell <- rbind(trend_cell, data.frame(id = i, Sen = NA, pvalue = NA, taxon = i))
       }else if(trend_i$pvalue < 0.05){
        trend_cell <- rbind(trend_cell, data.frame(id = j, trend_i))
        }else{
        trend_cell <- rbind(trend_cell, data.frame(id = i, Sen = NA, pvalue = NA, taxon = i))
      } 
   }
}
 
trend_cell <- trend_cell[-1, ]
trend_cell <- separate(trend_cell, id, c("Lon", "Lat"), sep = "_")
trend_cell$Lon <- as.numeric(trend_cell$Lon)
trend_cell$Lat <- as.numeric(trend_cell$Lat)
trend_cell <- na.omit(trend_cell)

write.csv(trend_cell, paste0(MFDir, "./trend_cell_traits_3.csv"), row.names = F)
```
```{r}
grid <- read.csv(paste0(dataDir, "/grid_projection.csv"))
trend_cell <- read.csv(paste0(MFDir, "./trend_cell_traits_3.csv"))

for (i in levels(as.factor(trend_cell$taxon))){
  trend_cell_i <- dplyr::filter(trend_cell, taxon == i)
  taxon_i <- sub("_", " ", i)
  
  sp <- ggplot(world) +
    geom_tile(data = grid, aes(x = Lon, y = Lat), fill = "#eee9e9") +
    geom_tile(data = trend_cell_i, aes(x = Lon, y = Lat, fill = Sen)) +
    geom_sf(data = world, fill = alpha("lightgrey", 1)) +
    coord_sf(xlim = c(min(df_predY$Lon), max(df_predY$Lon)), 
           ylim = c(min(df_predY$Lat), max(df_predY$Lat))) +
    theme_bw(150) +
    theme(panel.grid = element_blank(),
          legend.title = element_blank(), legend.position = "bottom",
          title = element_blank()) +
    scale_fill_gradient2(low = "#00c2c7", mid = "white", high = "#eb8c00", midpoint = 0, 
                         breaks = c(min(trend_cell_i$Sen, na.rm = T), 0, max(trend_cell_i$Sen, na.rm = T))) +
    guides(fill = guide_colourbar(barwidth = 90,
                                barheight = 2))
  
  try(dev.off())  
    png(paste0("./output/traits_phylo_3/", i,  "_trend_poly_traits_3.png"), height = 3000, width = 2100)
    print(sp)
    dev.off()
}
```

# HABITAT SUITABILITY CENTROIDES (LON/LAT - DEPTH)

## Calcul
```{r}
centroides <- data.frame(Year = NA, taxon = NA, depth = NA, Lat = NA, Lon = NA, geom = NA)
start <- 1997

for (i in 1:length(EpredY_list)){
  
  EpredY_i <- EpredY_list[[i]]
  EpredY_i <- data.frame(grid, EpredY_i)
  year_i <- start
  df_w_mean_i <- data.frame(Year = NA, taxon = NA, depth = NA, Lat = NA, Lon = NA, geom = NA)
  df_w_sd_i <- data.frame(Year = NA, taxon = NA, depth = NA, Lat = NA, Lon = NA, geom = NA)
  
  for(j in 11:ncol(EpredY_i)){
    taxon_j <- colnames(EpredY_i[j])
    
    w_mean_depth_j <- weighted.mean(EpredY_i$depth, EpredY_i[, j])
    w_mean_Lat_j <- weighted.mean(EpredY_i$Lat, EpredY_i[, j])
    w_mean_Lon_j <- weighted.mean(EpredY_i$Lon, EpredY_i[, j])
    
    df_w_mean_j <- data.frame(Year = year_i, taxon = taxon_j, depth = w_mean_depth_j, Lat = w_mean_Lat_j, Lon = w_mean_Lon_j, geom = "mean")
    df_w_mean_i <- rbind(df_w_mean_i, df_w_mean_j)
    
    w_sd_depth_j <- sqrt(wtd.var(EpredY_i$depth, EpredY_i[, j]))
    w_sd_Lat_j <- sqrt(wtd.var(EpredY_i$Lat, EpredY_i[, j]))
    w_sd_Lon_j <- sqrt(wtd.var(EpredY_i$Lon, EpredY_i[, j]))
    
    df_w_sd_j <- data.frame(Year = year_i, taxon = taxon_j, depth = w_sd_depth_j, Lat = w_sd_Lat_j, Lon = w_sd_Lon_j, geom = "standard_deviation")
    df_w_sd_i <- rbind(df_w_sd_i, df_w_sd_j)
  }
  
  centroides <- rbind(centroides, df_w_mean_i)
  centroides <- rbind(centroides, df_w_sd_i)
  
  centroides <- na.omit(centroides)
  start <- start + 1 
}

centroides <- pivot_longer(centroides, !c(Year, taxon, geom))
 

write.csv(centroides, file = paste0(MFDir, "/centroides_time_series_traits_3.csv"), row.names = F)
```

## Plots
```{r}
centroides <- read.csv(paste0(MFDir, "/centroides_time_series_traits_3.csv"))
centroides <- dplyr::filter(centroides, geom == "mean")

centroides <- relocate(centroides, Year, .after = taxon)
centroides <- centroides <- pivot_wider(centroides, values_from = value, names_from = name)

plot_names <- c('Amblyraja_radiata' = "Amblyraja radiata",
                'Leucoraja_naevus' = "Leucoraja naevus",
                'Mustelus_asterias' = "Mustelus asterias",
                'Raja_brachyura' = "Raja brachyura",
                'Raja_clavata' = "Raja clavata",
                'Raja_montagui' = "Raja montagui",
                'Scyliorhinus_canicula' = "Scyliorhinus canicula",
                'Scyliorhinus_stellaris' = "Scyliorhinus stellaris",
                "Squalus_acanthias" = 'Squalus acanthias')

ggplot() +
  geom_point(aes(x = Lon, y = Lat, col = Year), data = centroides) +
  facet_wrap(taxon ~ ., scales = "free", labeller = as_labeller(plot_names)) +
  theme_bw(15) +
  theme(panel.grid = element_blank()) +
  theme(legend.position = "bottom") +
  labs(x = "Longitude", y = "Latitude") +
  theme(plot.title = element_text(face = "italic")) +
  scale_colour_viridis_c()
```
```{r}
centroides <- read.csv(paste0(MFDir, "/centroides_time_series_traits_3.csv"))

plot_names <- c('depth' = "Depth (m)",
                'Lon' = "Longitude (degrees)",
                'Lat' = "Latitude (degrees))")
```
```{r}
centroides <- dplyr::filter(centroides, geom == "mean")
  
ggplot(aes(x = Year, y = value, col = taxon), data = centroides) +
  geom_line() +
  theme_bw(15)+
  labs(col = NULL, y = "Centroides", x = NULL) +
  theme(panel.grid = element_blank()) +
  facet_wrap(. ~ name, 
             scales = "free", 
             labeller = as_labeller(plot_names)) +
  theme(legend.position = "bottom") 
 
```

## Statistics
### Sen's estimates
```{r}
centroides <- read.csv(paste0(MFDir, "/centroides_time_series_traits_3.csv"))
centroides <- dplyr::filter(centroides, geom == "mean")

centroides <- relocate(centroides, Year, .after = taxon)
centroides <- centroides <- pivot_wider(centroides, values_from = value, names_from = name)
trend_stat <- data.frame(Sen = NA, pvalue = NA, taxon = NA, geo = NA)

for (i in levels(as.factor(centroides$taxon))){
  sp_i <- filter(centroides, taxon == i)
  
  try(test_lon_i <- sens.slope(sp_i$Lon, conf.level = 0.95))
  trend_i <- data.frame(Sen = test_lon_i$estimates, 
                        pvalue = test_lon_i$p.value, 
                        taxon = i, 
                        geo = "Lon")
 
  try(test_lat_i <- sens.slope(sp_i$Lat, conf.level = 0.95))
  trend_i <- rbind(trend_i, 
                   data.frame(Sen = test_lat_i$estimates, 
                              pvalue = test_lat_i$p.value,
                              taxon = i,
                              geo = "Lat"))
  
  try(test_depth_i <- sens.slope(sp_i$depth, conf.level = 0.95))
  trend_i <- rbind(trend_i, 
                   data.frame(Sen = test_depth_i$estimates,
                              pvalue = test_depth_i$p.value, 
                              taxon = i, 
                              geo = "depth"))
  
  trend_stat <- rbind(trend_stat, trend_i)
}

trend_stat <- trend_stat[-1, ]
write.csv(trend_stat, file = paste0(MFDir, "/centroides_time_series_stat_traits_3.csv"), row.names = F)

trend_stat <- filter(trend_stat, pvalue < 0.05)
trend_stat
```

### Plot - change

Dans le système géodésique WGS 84, au niveau de 54° de latitude nord, un degré de longitude équivaut à 65,576 km

```{r}
trend_stat <- read.csv(paste0(MFDir, "/centroides_time_series_stat_traits_3.csv"))
trend_stat <- dplyr::filter(trend_stat, taxon != "substrate_diversity")
trend_stat <- pivot_wider(trend_stat, names_from = geo, values_from = c(Sen, pvalue))


trend_stat$Sen_Lon_km <- (trend_stat$Sen_Lon) * 65.576
trend_stat$Sen_Lat_km <- (trend_stat$Sen_Lat) * 65.576

trend_stat$Lon_time_series <- (trend_stat$Sen_Lon_km) * 24
trend_stat$Lat_time_series <- (trend_stat$Sen_Lat_km) * 24
trend_stat$depth_time_series <- (trend_stat$Sen_depth) * 24
```
```{r}
Lon <- ggplot() +
  geom_col(aes (y = taxon, x = Sen_Lon_km, fill = taxon), data = trend_stat) +
  theme_bw(150) +
  theme(panel.grid = element_blank()) +
  scale_y_discrete(limits = c("Squalus_acanthias",
                              "Scyliorhinus_stellaris",
                              "Scyliorhinus_canicula",
                              "Raja_montagui",
                              "Raja_clavata",
                              "Raja_brachyura",
                              "Mustelus_asterias",
                              "Leucoraja_naevus",
                              "Amblyraja_radiata")) +
  scale_fill_manual(values = c("#0d2fb1ff","#999999ff", "#e69f00ff", "#56b4e9ff","#009e73ff", "#fff400ff","#0085f2ff","#e98377ff","#ff71fcff")) +
  guides(fill = "none")

try(dev.off())  
  png(paste0("./output/traits_phylo_3/Lon_change.png"), height = 1500, width = 3000)
    print(Lon)
    dev.off() 
```
```{r}
Lat <- ggplot() +
  geom_col(aes (y = taxon, x = Sen_Lat_km, fill = taxon), data = trend_stat) +
  theme_bw(150) +
  theme(panel.grid = element_blank())  +
  scale_y_discrete(limits = c("Squalus_acanthias",
                              "Scyliorhinus_stellaris",
                              "Scyliorhinus_canicula",
                              "Raja_montagui",
                              "Raja_clavata",
                              "Raja_brachyura",
                              "Mustelus_asterias",
                              "Leucoraja_naevus",
                              "Amblyraja_radiata")) +
  scale_fill_manual(values = c("#0d2fb1ff","#999999ff", "#e69f00ff", "#56b4e9ff","#009e73ff", "#fff400ff","#0085f2ff","#e98377ff","#ff71fcff"))+
  guides(fill = "none")

try(dev.off())  
  png(paste0("./output/traits_phylo_3/Lat_change.png"), height = 1500, width = 3000)
    print(Lat)
    dev.off() 
```
```{r}
depth <- ggplot() +
  geom_col(aes (y = taxon, x = Sen_depth, fill = taxon), data = trend_stat) +
  theme_bw(150) +
  theme(panel.grid = element_blank())  +
  scale_y_discrete(limits = c("Squalus_acanthias",
                              "Scyliorhinus_stellaris",
                              "Scyliorhinus_canicula",
                              "Raja_montagui",
                              "Raja_clavata",
                              "Raja_brachyura",
                              "Mustelus_asterias",
                              "Leucoraja_naevus",
                              "Amblyraja_radiata")) +
  scale_fill_manual(values = c("#0d2fb1ff","#999999ff", "#e69f00ff", "#56b4e9ff","#009e73ff", "#fff400ff","#0085f2ff","#e98377ff","#ff71fcff"))+
  guides(fill = "none")

try(dev.off())  
  png(paste0("./output/traits_phylo_3/depth_change.png"), height = 1500, width = 3000)
    print(depth)
    dev.off() 
```

# HABITAT SUITABILITY - FIRST/LAST DECILE

## Calcul
```{r}
centroides_deciles <- data.frame(Year = NA, taxon = NA, depth = NA, Lat = NA, Lon = NA, geom = NA)
start <- 1997

for (i in 1:length(EpredY_list)){
  
  EpredY_i <- EpredY_list[[i]]
  EpredY_i <- data.frame(grid, EpredY_i)
  year_i <- start
  
  df_w_mean_i <- data.frame(Year = NA, taxon = NA, depth = NA, Lat = NA, Lon = NA, geom = NA)
  
  for(j in 12:ncol(EpredY_i)){
    
    EpredY_i_j <- EpredY_i
    taxon_j <- colnames(EpredY_i_j[j])
    EpredY_i_j <- dplyr::filter(EpredY_i_j,  EpredY_i_j[[taxon_j]] != 0)
    
    EpredY_i_j$decile <- ntile(EpredY_i_j[j], 10)
    
    decile_1 <- dplyr::filter(EpredY_i_j, decile == 1)
    decile_10 <- dplyr::filter(EpredY_i_j, decile == 10)
    
    w_mean_depth_j_1 <- weighted.mean(decile_1$depth, decile_1[, j])
    w_mean_depth_j_10 <- weighted.mean(decile_10$depth, decile_10[, j])
    
    w_mean_Lat_j_1 <- weighted.mean(decile_1$Lat, decile_1[, j])
    w_mean_Lat_j_10 <- weighted.mean(decile_10$Lat, decile_10[, j])
    
    w_mean_Lon_j_1 <- weighted.mean(decile_1$Lon, decile_1[, j])
    w_mean_Lon_j_10 <- weighted.mean(decile_10$Lon, decile_10[, j])
    
    df_w_mean_j <- data.frame(Year = rep(year_i, 2),
                              taxon = rep(taxon_j, 2),
                              depth = c(w_mean_depth_j_1, w_mean_depth_j_10),
                              Lat = c(w_mean_Lat_j_1, w_mean_Lat_j_10),
                              Lon = c(w_mean_Lon_j_1, w_mean_Lon_j_10),
                              geom = c("first_mean", "last_mean"))
    
    df_w_mean_i <- rbind(df_w_mean_i, df_w_mean_j)
  }
  
  centroides_deciles <- rbind(centroides_deciles, df_w_mean_i)
  
  centroides_deciles <- na.omit(centroides_deciles)
  start <- start + 1 
}

centroides_deciles <- pivot_longer(centroides_deciles, !c(Year, taxon, geom))
 

write.csv(centroides_deciles, file = paste0(MFDir, "/centroides_deciles_time_series_traits_3.csv"), row.names = F)
```

## Plots
```{r}
centroides_deciles <- read.csv(paste0(MFDir, "/centroides_deciles_time_series_traits_3.csv"))

centroides_deciles <- relocate(centroides_deciles, Year, .after = taxon)

plot_names <- c('Amblyraja_radiata' = "Amblyraja radiata",
                'Leucoraja_naevus' = "Leucoraja naevus",
                'Mustelus_asterias' = "Mustelus asterias",
                'Raja_brachyura' = "Raja brachyura",
                'Raja_clavata' = "Raja clavata",
                'Raja_montagui' = "Raja montagui",
                'Scyliorhinus_canicula' = "Scyliorhinus canicula",
                'Scyliorhinus_stellaris' = "Scyliorhinus stellaris",
                "Squalus_acanthias" = 'Squalus acanthias')

ggplot() +
  geom_point(aes(x = Lon, y = Lat, col = Year), data = centroides_deciles) +
  facet_wrap(taxon ~ ., scales = "free", labeller = as_labeller(plot_names)) +
  theme_bw(15) +
  theme(panel.grid = element_blank()) +
  theme(legend.position = "bottom") +
  labs(x = "Longitude", y = "Latitude") +
  theme(plot.title = element_text(face = "italic")) +
  scale_colour_viridis_c()
```

## Statistics

### Sen's estimates
```{r}
centroides_deciles <- read.csv(paste0(MFDir, "/centroides_deciles_time_series_traits_3.csv"))

centroides_deciles <- relocate(centroides_deciles, Year, .after = taxon)
centroides_deciles <- pivot_wider(centroides_deciles, values_from = value, names_from = name)
trend_stat <- data.frame(Sen = NA, pvalue = NA, taxon = NA, geo = NA)

for (i in levels(as.factor(centroides$taxon))){
  sp_i <- dplyr::filter(centroides, taxon == i)
  
  first <- dplyr::filter(sp_i, geom == "first_mean")
  last <- dplyr::filter(sp_i, geom == "last_mean")
  
  try(test_lon_i <- sens.slope(first$Lon, conf.level = 0.95))
  trend_i <- data.frame(Sen = test_lon_i$estimates, 
                        pvalue = test_lon_i$p.value, 
                        taxon = i, 
                        geo = "first_lon")
  try(test_lon_i <- sens.slope(last$Lon, conf.level = 0.95))
  trend_i <- rbind(trend_i, 
                   data.frame(Sen = test_lon_i$estimates, 
                              pvalue = test_lon_i$p.value,
                              taxon = i,
                              geo = "last_lon"))
                        
  try(test_lat_i <- sens.slope(first$Lat, conf.level = 0.95))
  trend_i <- rbind(trend_i, 
                   data.frame(Sen = test_lat_i$estimates, 
                              pvalue = test_lat_i$p.value,
                              taxon = i,
                              geo = "first_lat"))
  try(test_lat_i <- sens.slope(last$Lat, conf.level = 0.95))
  trend_i <- rbind(trend_i, 
                   data.frame(Sen = test_lat_i$estimates, 
                              pvalue = test_lat_i$p.value,
                              taxon = i,
                              geo = "last_lat"))
  
  try(test_depth_i <- sens.slope(first$depth, conf.level = 0.95))
  trend_i <- rbind(trend_i, 
                   data.frame(Sen = test_depth_i$estimates,
                              pvalue = test_depth_i$p.value, 
                              taxon = i, 
                              geo = "first_depth"))
  try(test_depth_i <- sens.slope(last$depth, conf.level = 0.95))
  trend_i <- rbind(trend_i, 
                   data.frame(Sen = test_depth_i$estimates,
                              pvalue = test_depth_i$p.value, 
                              taxon = i, 
                              geo = "last_depth"))
  
  trend_stat <- rbind(trend_stat, trend_i)
}

trend_stat <- trend_stat[-1, ]
write.csv(trend_stat, file = paste0(MFDir, "/centroides_deciles_time_series_stat_traits_3.csv"), row.names = F)

trend_stat <- filter(trend_stat, pvalue < 0.05)
trend_stat
```


# NICHE SHAPE

## Function & data
```{r}
# Niche position functions
# polynomial responses of HMSC beta parameters

check_beta_trend <- function(LpredDeriv){
  pn<-dim(LpredDeriv)[2]
  response = matrix(NA,nrow = nrow(LpredDeriv),ncol = 1)
  for (i in 1:nrow(LpredDeriv)){
    if (LpredDeriv[i,1]>0 & LpredDeriv[i,pn]>0){   
      # derivative is positive at both ends -> clear case 1
      response[i] = 1 
    } else if (LpredDeriv[i,1]<0 & LpredDeriv[i,pn]<0) {
      # derivative is negative at both ends -> clear case 2
      response[i] = 2
    } else if (LpredDeriv[i,1]>0 & LpredDeriv[i,pn]<0) {
      # derivative changes from postitive to negative -> have to check in more detail
      
      if (sum(LpredDeriv[i,]>0)/pn<0.2){
        # derivative is positive over less than 20% of the x-gradient extent -> classify as case 2
        response[i] = 2
        
      } else if (sum(LpredDeriv[i,]>0)/pn>0.8){
        # derivative is positive over more than 80% of the x-gradient extent -> classify as case 1
        response[i] = 1
        
      } else {
        # derivative is postive or negative only over at most 60% of the x-gradient extent -> classify as case 3
        response[i] = 3
      }
      
    } else if (LpredDeriv[i,1]<0 & LpredDeriv[i,pn]>0) {
      # derivative changes from negative to positive -> have to check in more detail
      
      if (sum(LpredDeriv[i,]<0)/pn<0.2){
        # derivative is negative over less than 20% of the x-gradient extent -> classify as case 1
        response[i] = 1
        
      } else if (sum(LpredDeriv[i,]<0)/pn>0.8){
        # derivative is negative over more than 80% of the x-gradient extent -> classify as case 2
        response[i] = 2
        
      } else {response[i] = 3 
      # derivative is positive or negative only over at most 60% of the x-gradient extent 
      
      }
    }
    
  }
  # calculate the probabilities for the cases
  Pr1 = sum(response==1,na.rm=T)/length(response) #positive; derivative is positive over more than 90%
  Pr2 = sum(response==2,na.rm=T)/length(response) #negative, derivative is negative over more than 90%
  Pr3 = sum(response==3,na.rm=T)/length(response) #change; derivative is positive or negative only over at most 60% 
  
  return(c(Pr1,Pr2,Pr3))}
```
```{r}
load(paste0(ModelDir, "/pa_traits_phylo_probit_thin_3.RData"))
envt <- read.csv(paste0(dataDir, "/hindcast/envt_hindcast_traits.csv"))
envt <- dplyr::select(envt, Year, max_SST_summer, max_ph_summer)
```

## Calcul
```{r}
mpost <- convertToCodaObject(m)
Beta <- as.matrix(mpost$Beta)
XData <- m$XData
XData <- merge(XData, envt)
spn <- ncol(m$Y[])
poly$taxon <- colnames(m$Y[])

temp_pred <- as.vector(seq(min(XData$max_SST_summer), 
                          max(XData$max_SST_summer), length = 100))
ph_pred <- as.vector(seq(min(XData$max_ph_summer),
                        max(XData$max_ph_summer),length = 100))

for (sp in 0:(spn-1)){
  sp1 <- Beta[, (sp * 9 + 1) : (sp * 9 + 9)]
          
  LpredDeriv <- sp1[, 4] + 2 * sp1[, 5] %*% t(temp_pred)
  res_temp <- rbind(res_temp, check_beta_trend(LpredDeriv))
          
  LpredDeriv = sp1[, 6] + 2 * sp1[, 7] %*% t(ph_pred)
  res_ph <- rbind(res_ph, check_beta_trend(LpredDeriv))
          
  colnames(res_temp) <- c('pos_res_temp','neg_res_temp','change_res_temp')
  colnames(res_ph) <- c('pos_res_ph','neg_res_ph','change_res_ph')
        
  poly <- data.frame(res_temp, res_ph)
}

poly$taxon <- colnames(m$Y[])

write.csv(poly, paste0(MFDir, "/poly_trend_stat_traits_3.csv"), row.names = F)
```

### 1997-2009
```{r}
SXData <- dplyr::filter(XData, Year < 2010)

res_temp <- c()
res_ph <- c()

temp_pred <- as.vector(seq(min(SXData$max_SST_summer), 
                          max(SXData$max_SST_summer), length = 100))
ph_pred <- as.vector(seq(min(SXData$max_ph_summer),
                        max(SXData$max_ph_summer),length = 100))

for (sp in 0:(spn-1)){
  sp1 <- Beta[, (sp * 9 + 1) : (sp * 9 + 9)]
          
  LpredDeriv <- sp1[, 4] + 2 * sp1[, 5] %*% t(temp_pred)
  res_temp <- rbind(res_temp, check_beta_trend(LpredDeriv))
          
  LpredDeriv = sp1[, 6] + 2 * sp1[, 7] %*% t(ph_pred)
  res_ph <- rbind(res_ph, check_beta_trend(LpredDeriv))
          
  colnames(res_temp) <- c('pos_res_temp','neg_res_temp','change_res_temp')
  colnames(res_ph) <- c('pos_res_ph','neg_res_ph','change_res_ph')
        
  poly <- data.frame(res_temp, res_ph)
}

write.csv(poly, paste0(MFDir, "/poly_trend_stat_traits_3_1997_2009.csv"), row.names = F)
```

### 2010-2020
```{r}
EXData <- dplyr::filter(XData, Year >= 2010)

res_temp <- c()
res_ph <- c()

temp_pred <- as.vector(seq(min(EXData$max_SST_summer), 
                          max(EXData$max_SST_summer), length = 100))
ph_pred <- as.vector(seq(min(EXData$max_ph_summer),
                        max(EXData$max_ph_summer),length = 100))

for (sp in 0:(spn-1)){
  sp1 <- Beta[, (sp * 9 + 1) : (sp * 9 + 9)]
          
  LpredDeriv <- sp1[, 4] + 2 * sp1[, 5] %*% t(temp_pred)
  res_temp <- rbind(res_temp, check_beta_trend(LpredDeriv))
          
  LpredDeriv = sp1[, 6] + 2 * sp1[, 7] %*% t(ph_pred)
  res_ph <- rbind(res_ph, check_beta_trend(LpredDeriv))
          
  colnames(res_temp) <- c('pos_res_temp','neg_res_temp','change_res_temp')
  colnames(res_ph) <- c('pos_res_ph','neg_res_ph','change_res_ph')
        
  poly <- data.frame(res_temp, res_ph)
}

write.csv(poly, paste0(MFDir, "/poly_trend_stat_traits_3_2010_2020.csv"), row.names = F)
```

## Plot
```{r}
area_stat <- read.csv(paste0(MFDir, "/weighted_area_trend_stat_traits_3.csv"))
#poly_stat <- read.csv(paste0(MFDir, "/poly_trend_stat_traits_3.csv"))

stat <- area_stat[, 3:4]
colnames(stat) <- c("taxon", "area.size")
```

### Niche positioning

#### 1997-2009
```{r}
poly_stat <- read.csv(paste0(MFDir, "/poly_trend_stat_traits_3_2010_2020.csv"))
```
```{r}
poly_stat$pos_res_temp[poly_stat$pos_res_temp >= 0.8] <- "Increasing"
poly_stat$neg_res_temp[poly_stat$neg_res_temp >= 0.8] <- "Decreasing"

poly_stat$pos_res_ph[poly_stat$pos_res_ph >= 0.8] <- "Increasing"
poly_stat$neg_res_ph[poly_stat$neg_res_ph >= 0.8] <- "Decreasing"

poly_stat$change_res_ph[poly_stat$change_res_ph >= 0.8] <- "Bell-shaped"
poly_stat$change_res_temp[poly_stat$change_res_temp >= 0.8] <- "Bell-shaped"

poly_stat
```

With traits
```{r}
poly_stat <- data.frame(taxon = poly_stat$taxon, 
                        temp = c("Bell-shaped", "Decreasing", "Bell-shaped", "Decreasing", "Bell-shaped", "Decreasing", "Decreasing", "Bell-shaped", "Decreasing"),
                        ph = c("Bell-shaped", "Bell-shaped", "Bell-shaped", "Bell-shaped", "Bell-shaped", "Bell-shaped", "Bell-shaped", "Decreasing", "Bell-shaped"))

stat <- merge(stat, poly_stat, by = "taxon")
```

```{r}
cbPalette <- c("#0d2fb1", "#999999", "#E69F00", "#56B4E9", "#009E73", "#fff400", "#0085f2","#e85645", "#ff71fc")

plot <- ggplot(data = stat, aes(axis1 = ph, axis2 = temp, axis3 = area.size)) +
  geom_alluvium(aes(fill = taxon), alpha = 5/7) +
  geom_stratum(fill = "white") + 
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_bw(18) +
  theme(panel.grid = element_blank()) +
  theme(axis.title = element_blank())  +
  theme (legend.title = element_blank()) +
  theme(legend.position = "bottom") +
  scale_x_discrete(limits= c("Species response\nto pH", "Species response\nto temperature", "Suitable habitat area")) +
  scale_y_discrete(breaks = NULL)  +
  scale_fill_manual(values = cbPalette,
                    labels = c("A. radiata", "L. naevus", "M. asterias", "R. brachyura", "R. clavata", "R. montagui", "S. canicula", "S. stellaris", "S. acanthias")) +
  theme(legend.text = element_text(face = "italic"))

try(dev.off())  
  png("./output/traits_phylo_3/species_responses_poly_traits_3.png", width = 600)
    print(plot)
    dev.off()
```



